Exoplanet population inference and the abundance \\ of Earth analogs from noisy, incomplete catalogs

### 2014-06-11 document/ms.tex
a few typos etc

\section{Introduction}

@@

\citet{petigura} recently published an exoplanet population analysis based on
an independent analysis of the \kepler\ light curves for 42,557 Sun-like
stars.
This study was especially novel because the authors developed an independent
planet search pipeline (\terra; \citealt{petigura-a}) and determined the
detection efficiency of their analysis empirically by injecting synthetic
signals into real light curves measured by \kepler.
The occurrence rate function determined by \citet{petigura} agreed
qualitatively with previous studies of small planets orbiting Sun-like stars
(\citealt{dong}).
In particular, both papers describe a ``flattening'' rate function (in
logarithmic radius) for planets around Earth's radius.
Even though no Earth analogs were discovered in their search, \citet{petigura}
used the small candidates that they did find to place an extrapolated
constraint on the frequency of Earth-like exoplanets, assuming a flat
occurrence rate density in logarithmic period.

@@

There are two qualitatively different methods that are commonly used to infer
the occurrence rate density from a catalog and a detection efficiency
specification.
The first is an intuitive method that we will refer to as
``inverse-detection-efficiency'' and the second is based on the likelihood
function of the catalog given a parametric rate density.
The inverse-detection-efficiency method involves making a histogram of the
objects in the catalog where each point is weighted by its inverse detection
probability.
This method is very popular in the literature (\citealt{howard, dong,
dressing, swift, petigura}) even though it is not motivated probabilistically.
The alternative likelihood method models the catalog as a Poisson realization
of the \emph{observable} rate density of exoplanets taking the survey
detection efficiencies and transit probabilities into account.
This technique has been used to constrain parametric models---a broken power
law, for example---for the occurrence rate density (\citealt{tabachnik,
youdin, dong}).
A result based on an optimization of a likelihood function will---under any
stated assumptions---saturate the Cram\'er--Rao bound and give an estimator of
the population with the lowest possible variance.
As we show in \sect{model}, if we model the occurrence rate density as a
piecewise constant function, the inverse-detection-efficiency method can be
derived as a special case of the likelihood method with a specific set of
assumptions.

@@

For the purposes of this \paper, we make some strong assumptions, although we
argue that they are weaker than the implicit assumptions in previous
studies.
None of these assumptions is necessary for the validity of our general method
but they do simplify the specific procedures we employ.
Specifically, we assume that
\begin{itemize}

@@

We must emphasize one very important consequence of our assumptions.
We assume that the catalog of exoplanet candidates is only missing planets
with probabilities given by the empirical detection efficiency.
In detail this must be false; the detection efficiency we use
doesn't take into account the fact that the catalog doesn't include multiple
transiting systems.
A large fraction of the transiting planets discovered by the \kepler\ transit
search pipeline are members of multiple transiting systems (see
\citealt{lissauer}, for example).
Since \citet{petigura} only detected at most one planet per system, their
catalog is actually a list of planet candidates \emph{without a more
detectable companion}.
The global effects of this selection is not trivial and an in-depth discussion
is beyond the scope of this \paper\ but all of the results should be
interpreted with this caveat in mind.

@@

Any inference about the rate density parameters can be formally addressed by
evaluating the \emph{marginalized likelihood function}
\begin{eqnarray}\eqlabel{crazylike}
p(\{\data_k\}\,|\,\ratepars) &=&
    \int p(\{\data_k\}\,|\,\{\entry_k\})
    \,p(\{\entry_k\}\,|\,\ratepars)
    \dd\{\entry_k\}
\end{eqnarray}
where, as we mentioned above, $\{\data_k\}$ is the set of all light curves,
one light curve $\data_k$ per target $k$, \ratepars\ is the vector of
parameters describing the population occurrence rate density
$\rate_\ratepars(\entry)$ and $\entry_k$ is the vector of physical parameters
describing the planetary system (orbital periods, radius ratios, stellar
radius, \etc) around target $k$.
In this equation, our only assumption is that the datasets depend on the
rate density of exoplanets only through the catalog $\{\entry_k\}$.
In our case, this assumption qualitatively means that the signals found in the
light curves depend only on the actual planet properties and not on the
distributions from which they are drawn.
It is worth emphasizing that---as we will discuss further below---the catalog
only provides \emph{probabilistic constraints} on $\{\entry_k\}$; not perfect
delta-function measurements.

@@

The constraints in \eq{crazypost} can always be---and often are---propagated
as a list of $N$ samples $\{\entry_k\}^{(n)}$ from the posterior
\begin{eqnarray}\eqlabel{samples}
\{\entry_k\}^{(n)} &\sim& p(\{\entry_k\}\,|\,\{\data_k\},\,\interim) \quad.
\end{eqnarray}
We can use these samples and the Monte Carlo integral approximation to
estimate the marginalized likelihood from \eq{simplemarglike}---up to an
irrelevant constant---as
\begin{eqnarray}\eqlabel{importance}
p(\{\data_k\}\,|\,\ratepars) &\approx&
    \frac{\mathcal{Z}}{N} \sum_{n=1}^N
    \frac{p(\{\entry_k\}^{(n)}\,|\,\ratepars)}
         {p(\{\entry_k\}^{(n)}\,|\,\interim)}
\end{eqnarray}
where the constant $\mathcal{Z} = p(\{\data_k\}\,|\,\interim)$ is not a
function of the parameters.
This is very efficient to compute as long as an evaluation of
$p(\{\entry_k\}\,|\,\ratepars)$ is not very expensive.
That being said, this could be a high variance estimator, depending on the
number of independent samples $N$ and the initial choice of
$p(\{\entry_k\}\,|\,\interim)$.
Additionally, the support of $p(\{\entry_k\}\,|\,\ratepars)$ in $\{\entry_k\}$
space is restricted to be narrower than that of
$p(\{\entry_k\}\,|\,\interim)$.
Besides this caveat, in the limit of infinite samples, the approximation in
\eq{importance} becomes exact.
\Eq{importance} is the \emph{importance sampling approximation} to the
integral in \eq{simplemarglike} where the trial density is the posterior
probability for the catalog measurements.

@@

In the inference procedure described above, we only discussed an abstract
occurrence rate density.
To make it more concrete, we'll model the catalog as a Poisson draw from the
\emph{observable} rate density $\obs{\rate}_\ratepars$.
This leads to the previously known result (see \citealt{tabachnik,youdin} for
some of the examples from the exoplanet literature)
\begin{eqnarray}\eqlabel{poisson-like}
p(\{\entry_k\}\,|\,\ratepars) &=&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K \obs{\rate}_\ratepars (\entry_k)\quad.
\end{eqnarray}
In this equation, the integral in the normalization term is the expected
number of observable exoplanets in the sample.
A further assumption that we have made in \eq{poisson-like} is that each
object is an independent sample from the inhomogeneous Poisson process set by
the rate density.
This assumption---though it is often made---cannot be correct when we consider
systems of multiple planets.
It is possible to relax this assumption (see \citealt{tremaine} for a related
technique) but this generalization is beyond the scope of the current \paper.

The main thing to note here is that $\obs{\rate}_\ratepars$ is the rate
density of exoplanets that you would expect to observe taking into account the
geometric transit probability and any other detection efficiencies.
In practice, we can model the observable rate density as
\begin{eqnarray}\eqlabel{obs-rate}
\obs{\rate}_\ratepars(\entry) &=&
    \completeness(\entry)\,\rate_\ratepars(\entry)
\end{eqnarray}
where $\completeness(\entry)$ is the detection efficiency (including transit
probability) at \entry\ and $\rate_\ratepars(\entry)$ is the object that we
want to infer: the \True\ occurrence rate density.
Up to this point, we haven't discussed any specific functional form for
$\rate_\ratepars(\entry)$ and all of this derivation is equally applicable
whether we model the rate density as (for example) a broken power law or
a histogram.

@@

For the results in this \paper, we will assume that the completeness function
$\completeness(\entry)$ is known empirically but that is not a requirement for
the validity of this method.
Instead, we could use a functional form for it and even infer its parameters
along with the parameters of the rate density.

@@

There is a very efficient algorithm called elliptical slice sampling (ESS;
\citealt{ess, ess-hyper}) for sampling the step heights \ratepars\ from the
density in \eq{joint-posterior}.
In practice, for problems with this specific structure, ESS outperforms more
traditional MCMC methods commonly employed in astrophysics
(e.g.,~\citealt{emcee}).
Our implementation is adapted from Jo Bovy's BSD licensed ESS
code\footnote{\url{https://github.com/jobovy/bovy_mcmc/blob/master/bovy_mcmc/%
elliptical_slice.py}}.
To simultaneously marginalize over the hyperparameter choice, we use the
Metropolis--Hastings update from Algorithm 1 in \citet{ess-hyper}.
We tune the Metropolis--Hastings proposal by hand until we get an acceptance
fraction of $\sim0.2-0.4$ for the hyperparameters.

For all the results below, we run a Markov chain with $10^6$ steps for the
heights and update the hyperparameters every 10 steps.
We only keep the final $2 \times 10^5$ steps and discard the earlier samples
as burn-in.
This results in approximately 4000 independent posterior samples estimated
using the empirical integrated autocorrelation time of the chain
(\citealt{goodman}).
These samples provide an approximation to the marginalized probability
distribution for \ratepars.

@@

The detection efficiency also depends on the geometric transit probability
$R_\star/a$.
Since we are modeling the distribution in the period--radius plane, we need to
compute the transit probability marginalized over stellar radius and mass.
This marginalized distribution scales only with the period of the orbit as
$\propto \period^{-2/3}$.
In theory, this marginalization should be over the \True\ distribution of
these parameters in the selected stellar catalog but we'll approximate it by
the empirical distribution; a reasonable approximation given the size of the
dataset.
At a period of 10 days\footnote{This period is chosen arbitrarily because the
power law only needs to be normalized at one point.}, the median transit
probability in the selected sample of stars is $5.061\%$ so we model the
transit probability\footnote{We are using the letter $Q$ to indicate
probabilities since we are already using $P$ to mean period.} as a function of
period as
\begin{eqnarray}\eqlabel{transitprob}
\transitprob (\period) &=&
    0.05061\,\left[\frac{\period}{10\,\mathrm{days}}\right]^{-2/3} \quad.
\end{eqnarray}
This expression is clearly only valid for $\period \gtrsim 1.4\,\mathrm{days}$
but the dataset that we are using (\citealt{petigura}) explicitly only
includes periods longer than five days so this is not a problem.
We're using the \emph{median} transit probability (instead of the mean)
because it is a more robust estimator in the presence of outliers but in our
experiments, the results do not seem to be very sensitive to this choice.

@@

In order to get a feeling for the constraints provided by our method and to
explore any biases introduced by ignoring the observational uncertainties, we
start by ``observing'' a few synthetic catalogs from qualitatively different
known occurrence rate density functions.
For each of these simulations, we'll take the completeness function computed
by \citet{petigura} as given.
In general, \eq{poisson-like} can be sampled using a procedure called thinning
(\citealt{poisson}) but for our purposes, we'll simply consider a piecewise
constant rate density evaluated on a fine grid in log-period and log-radius.
For this discrete function, the generative procedure is simple;
\begin{enumerate}
{\item loop over each grid cell $i$,}
{\item draw Poisson random integer $K_i\sim\mathrm{Poissson}(\obs{\rate}_i)$
with the observable rate density in the cell, and}
{\item distribute $K_i$ catalog entries in the cell randomly.}
\end{enumerate}
We then choose fractional observational uncertainties on the radii from the
\citet{petigura} catalog and apply them to the true catalog as Gaussian noise.

@@

Since no Earth analogs have been found, any constraints on this density must
be extrapolated from the existing observations.
This is generally done by assuming a functional form for the occurrence rate
density, constraining it using the observed candidates and extrapolating.
All published extrapolations are based on rigid models of the occurrence rate
density (for example, a power law) fit to the catalog and evaluated at the
location of Earth (\citealt{catanzarite, traub}).
\citet{petigura} used their catalog of planet candidates to constrain the rate
of Earth analogs in a specific period--radius bin assuming an extremely rigid
model: \emph{flat in logarithmic period}.
These results are sensitive to the choice of extrapolation function and the
specific definition of ``Earth analog''.

@@

Figures~\figref{smooth-rate} and~\figref{simulation-rate} compare our results
and the results of the \citet{petigura} extrapolation procedure when applied
to the synthetic catalogs.
Since these catalogs were simulated from a known population model, we know the
true value of \gammaearth\ and it is indicated in the Figures with a vertical
gray line.
In both cases, our method returns a less precise but more accurate result for
the rate density and the error bars given by the functional extrapolation
are overly optimistic.
One major effect that leads to this bias is that the period distribution is
not flat.
Restricting the result to only include uniform models is equivalent to
applying an extremely informative prior that doesn't have enough freedom to
capture the complexity of the problem.
As a result, the posterior constraints on \gammaearth\ are dominated by this
prior choice and the resulting uncertainties are much smaller than they should
be.

@@

The equivalent Figures for the radius distribution are shown in
Figures~\figref{radius} and~\figref{linear-radius}.
\Fig{radius} shows the log-radius occurrence rate density integrated over
various logarithmic bins in period.
The distributions in each period bin are qualitatively consistent; the
rate density is dominated by small planets (around two Earth radii) with
potential ``features'' near $\radius\sim3\radius_\oplus$ and $\radius\sim
10\radius_\oplus$.
These features appear in every period bin and they were also detected---using
a completely different dataset and technique---by \citet{dong} and a similar
result is visible in the occurrence rate determined by \citet[][their Figure
7]{fressin-fp} at low signal-to-noise.
\Fig{linear-radius} shows the same result but presented as a function of
linear radius.
In these coordinates, the rate density in a single bin is no longer
uniform; instead, scales as inverse radius.

@@

Our inferred rate density of Earth analogs (\eqalt{ge-result}) is not
consistent with the previously published result even though we use
\emph{exactly the same dataset}.
This inconsistency is due to the different assumptions made.
In particular, for their analysis, \citet{petigura} asserted that the period
distribution for small planets on long orbits is flat in logarithmic period.
If we enforce the same constraint, we reproduce their result but we relax this
assumption, allowing the data to constrain the period distribution, we find
evidence for a \emph{decreasing} period distribution as a function of
logarithmic period (see \fig{period}).

@@

We have developed a hierarchical probabilistic framework for inferring the
population of exoplanets based on noisy incomplete catalogs.
This method incorporates systematic treatment of observational uncertainties
and detection efficiency.
One major benefit of this framework is that it provides the best possible
probabilistic measurements of the population under the stated assumptions.
After demonstrating the validity of our method on two qualitatively different
synthetic exoplanet catalogs, we run our inference on a published catalog of
small exoplanet candidates orbiting Sun-like stars (\citealt{petigura}) to
determine the occurrence rate density of exoplanets as a function of period
and radius.
We extrapolate this measurement to the location of Earth and constrain the
rate density of Earth analogs with large error bars.
In order to perform this extrapolation, we don't assume a specific functional
form for the rate density.
Instead, we only assume that it is a smooth function of logarithmic period and
radius.

@@

Our extrapolation of the rate density to the location of Earth is more general
and conservative than any previously published method.
Using this method, we find a rate density of Earth analogs that is
inconsistent with the result published by \citet{petigura}.
This discrepancy can be largely attributed to the rigidity of the assumptions
about the period distribution.
Our extrapolation is less confident than previous measurements.
Again, this difference is due to the fact that we allow a much more flexible
extrapolation function.
This is another illustration that, against the standard data analysis
folklore, the correct use of flexible models is \emph{conservative}.

@@

\item {\bf Conditional independence}\quad
We assume that every object in the catalog was a conditionally independent
draw from the observable occurrence rate density.
This is probably a bad assumption when applying this method to a different
catalog where multiple transiting systems are included.
In practice, the best first step towards relaxing this assumption is probably
to follow \citet{tremaine} and assume that the mutual inclination distribution
is the only source of conditional dependence between planets.
For this \paper, the assumption of conditional independence is justified
because the dataset explicitly includes only systems with a single transiting
exoplanet.

@@

\item {\bf Given empirical detection efficiency}\quad
\citet{petigura} determined the end-to-end detection efficiency of their
planet detection pipeline as a function of \True\ period and radius by
injecting synthetic signals into real light curves and testing recovery.
We used these simulations as an exact representation of the detection
efficiency of the catalog but there are several missing components.
The biggest effect is probably the fact that this formulation doesn't include
the fact that only the \emph{most detectable signal} in each light curve was
included in the catalog.
This bias will have the largest effect in the parts of parameter space where
the baseline detection efficiency is lowest: at long periods and small radius.
As a result, our inferences (and the results from \citealt{petigura}) about
the occurrence rate of small planets on long periods should be seen as
\emph{lower limits}.
In detail there is another limitation due to the fact that the stellar
parameters are only known noisily and the transit light curve only constrains
the radius ratio.
This means that the marginalized detection efficiency should be measured as a
function of radius ratio and the interpretation in terms of \True\ radius is
only approximately correct.
Given the size of the dataset and the number of injection simulations, this
effect should be small.

@@

Under the assumptions that we have made here, our inference of the occurrence
rate density of exoplanets places a probabilistic constraint on the number of
transiting Earth analogs in the existing \kepler\ dataset.
If we adopt the definition of ``Earth-like'' from \citet[][$200 \le
\period/\mathrm{day} < 400$ and $1 \le \radius/\radius_\oplus < 2$]{petigura},
and integrate the product inferred rate density function and the geometric
transit probability (\eqalt{transitprob}) over this bin, we find that there
should be
\begin{eqnarray}\eqlabel{ntransit}
N_{\oplus,\,\mathrm{transiting}} &=& 9.2_{-4.0}^{+5.9}
\end{eqnarray}
Earth-like exoplanets transiting the stars in the sample of Sun-like stars
chosen by \citet{petigura}.
This is an exciting result because it means that, if we can improve the
sensitivity of exoplanet search pipelines to small planets orbiting on long
periods then we should find some Earth analogs in the existing data.
Furthermore, because of the treatment of multiple transiting systems in the
catalog, the number in \eq{ntransit} is probably a lower limit on the expected
number of Earth-like exoplanets transiting Sun-like stars!

>>>>

\section{Introduction}
\sectlabel{intro}

@@

\citet{petigura} recently published an exoplanet population analysis based on
an independent study of the \kepler\ light curves for 42,557 Sun-like stars.
This study was especially novel because the authors developed their own
planet search pipeline (\terra; \citealt{petigura-a}) and determined the
detection efficiency of their analysis empirically by injecting synthetic
signals into real light curves measured by \kepler.
The occurrence rate function determined by \citet{petigura} agrees
qualitatively with previous studies of small planets orbiting Sun-like stars
(\citealt{dong}).
In particular, both papers describe a ``flattening'' rate function (in
logarithmic radius) for planets around Earth's radius.
Even though no Earth analogs were discovered in their search, \citet{petigura}
used the small candidates that they did find to place an extrapolated
constraint on the frequency of Earth-like exoplanets, assuming a flat
occurrence rate density in logarithmic period.

@@

There are two qualitatively different methods that are commonly used to infer
the occurrence rate density from a catalog and a detection efficiency
specification.
The first is an intuitive method that we will refer to as
``inverse-detection-efficiency'' and the second is based on the likelihood
function of the catalog given a parametric rate density.
The inverse-detection-efficiency method involves making a histogram of the
objects in the catalog where each point is weighted by its inverse detection
probability.
This method is very popular in the literature (\citealt{howard, dong,
dressing, swift, petigura}) even though it is not motivated probabilistically.
The alternative likelihood method models the catalog as a Poisson realization
of the \emph{observable} rate density of exoplanets taking the survey
detection efficiencies and transit probabilities into account.
This technique has been used to constrain parametric models---a broken power
law, for example---for the occurrence rate density (\citealt{tabachnik,
youdin, dong}).
As we show in \sect{model}, if we model the occurrence rate density as a
piecewise constant function, the inverse-detection-efficiency method can be
derived as a special case of the likelihood method with a specific set of
assumptions.

@@

For the purposes of this \paper, we make some strong assumptions, although we
argue that they are weaker than the implicit assumptions in previous
studies.
None of these assumptions is necessary for the validity of our general method
but they do simplify the specific procedures we employ.
We assume that
\begin{itemize}

@@

We must emphasize one very important consequence of our assumptions.
We assume that the catalog of exoplanet candidates is only missing planets
with probabilities given by the empirical detection efficiency.
In detail this must be false; the detection efficiency we use
doesn't take into account the fact that the catalog doesn't include multiple
transiting systems.
A large fraction of the transiting planets discovered by the \kepler\ transit
search pipeline are members of multiple transiting systems (see
\citealt{lissauer}, for example).
Since \citet{petigura} only detected at most one planet per system, their
catalog is actually a list of planet candidates \emph{without a more
detectable companion}.
The global effects of this selection are not trivial and an in-depth
discussion is beyond the scope of this \paper\ but all of the results should
be interpreted with this caveat in mind.

@@

Any inference about the rate density parameters can be formally addressed by
evaluating the \emph{marginalized likelihood function}
\begin{eqnarray}\eqlabel{crazylike}
p(\{\data_k\}\,|\,\ratepars) &=&
    \int p(\{\data_k\}\,|\,\{\entry_k\})
    \,p(\{\entry_k\}\,|\,\ratepars)
    \dd\{\entry_k\}
\end{eqnarray}
where, as we mentioned above, $\{\data_k\}$ is the set of all light curves,
one light curve $\data_k$ per target $k$, \ratepars\ is the vector of
parameters describing the population occurrence rate density
$\rate_\ratepars(\entry)$ and $\entry_k$ is the vector of physical parameters
describing the planetary system (orbital periods, radius ratios, stellar
radius, \etc) around target $k$.
In this equation, our only assumption is that the datasets depend on the
rate density of exoplanets only through the catalog $\{\entry_k\}$.
In our case, this assumption qualitatively means that the signals found in the
light curves depend only on the actual properties of the planet and star, and
not on the distributions from which they are drawn.
It is worth emphasizing that---as we will discuss further below---the catalog
only provides \emph{probabilistic constraints} on $\{\entry_k\}$; not perfect
delta-function measurements.

@@

The constraints in \eq{crazypost} can always be---and often are---propagated
as a list of $N$ samples $\{\entry_k\}^{(n)}$ from the posterior
\begin{eqnarray}\eqlabel{samples}
\{\entry_k\}^{(n)} &\sim& p(\{\entry_k\}\,|\,\{\data_k\},\,\interim) \quad.
\end{eqnarray}
We can use these samples and the Monte Carlo integral approximation to
estimate the marginalized likelihood from \eq{simplemarglike}---up to an
irrelevant constant---as
\begin{eqnarray}\eqlabel{importance}
p(\{\data_k\}\,|\,\ratepars) &\approx&
    \frac{Z_\interim}{N} \sum_{n=1}^N
    \frac{p(\{\entry_k\}^{(n)}\,|\,\ratepars)}
         {p(\{\entry_k\}^{(n)}\,|\,\interim)}
\end{eqnarray}
where the constant $Z_\interim = p(\{\data_k\}\,|\,\interim)$ is not a
function of the parameters \ratepars.
This is very efficient to compute as long as an evaluation of
$p(\{\entry_k\}\,|\,\ratepars)$ is not expensive.
That being said, \eq{importance} could be a high variance estimator of
\eq{simplemarglike}, depending on the number of independent samples $N$ and
the initial choice of $p(\{\entry_k\}\,|\,\interim)$.
Additionally, the support of $p(\{\entry_k\}\,|\,\ratepars)$ in $\{\entry_k\}$
space is restricted to be narrower than that of
$p(\{\entry_k\}\,|\,\interim)$.
Besides this caveat, in the limit of infinite samples, the approximation in
\eq{importance} becomes exact.
\Eq{importance} is the \emph{importance sampling approximation} to the
integral in \eq{simplemarglike} where the trial density is the posterior
probability for the catalog measurements.

@@

In the inference procedure described above, we only discussed an abstract
occurrence rate density.
To make it more concrete, we'll model the catalog as a draw from the
inhomogeneous Poisson process set by the \emph{observable} rate density
$\obs{\rate}_\ratepars$.
This leads to the previously known result (see \citealt{tabachnik,youdin} for
some of the examples from the exoplanet literature)
\begin{eqnarray}\eqlabel{poisson-like}
p(\{\entry_k\}\,|\,\ratepars) &=&
    \exp\left(-\int \obs{\rate}_\ratepars (\entry) \dd\entry\right) \,
    \prod_{k=1}^K \obs{\rate}_\ratepars (\entry_k)\quad.
\end{eqnarray}
In this equation, the integral in the normalization term is the expected
number of observable exoplanets in the sample.

The main thing to note here is that $\obs{\rate}_\ratepars$ is the rate
density of exoplanets that you would expect to observe taking into account the
geometric transit probability and any other detection efficiencies.
In practice, we can model the observable rate density as
\begin{eqnarray}\eqlabel{obs-rate}
\obs{\rate}_\ratepars(\entry) &=&
    \completeness(\entry)\,\rate_\ratepars(\entry)
\end{eqnarray}
where $\completeness(\entry)$ is the detection efficiency (including transit
probability) at \entry\ and $\rate_\ratepars(\entry)$ is the object that we
want to infer: the \True\ occurrence rate density.
Up to this point, we haven't discussed any specific functional form for
$\rate_\ratepars(\entry)$ and all of this derivation is equally applicable
whether we model the rate density as, for example, a broken power law or
a histogram.

@@

For the results in this \paper, we will assume that the completeness function
$\completeness(\entry)$ is known empirically on a grid in period and radius
but that is not a requirement for the validity of this method.
Instead, we could use a functional form for the completeness and even infer
its parameters along with the parameters of the rate density.

@@

There is an efficient algorithm called elliptical slice sampling (ESS;
\citealt{ess, ess-hyper}) for sampling the step heights \ratepars\ from the
density in \eq{joint-posterior}.
In practice, for problems with this specific structure, ESS outperforms more
traditional MCMC methods commonly employed in astrophysics
(e.g.,~\citealt{emcee}).
Our implementation is adapted from Jo Bovy's BSD licensed ESS
code\footnote{\url{https://github.com/jobovy/bovy_mcmc/blob/master/bovy_mcmc/%
elliptical_slice.py}}.
To simultaneously marginalize over the hyperparameter choice, we use the
Metropolis--Hastings update from Algorithm 1 in \citet{ess-hyper}.
We tune the Metropolis--Hastings proposal by hand until we get an acceptance
fraction of $\sim0.2-0.4$ for the hyperparameters.

For all the results below, we run a Markov chain with $10^6$ steps for the
heights and update the hyperparameters every 10 steps.
We only keep the final $2 \times 10^5$ steps and discard the earlier samples
as burn-in.
By estimating the empirical integrated autocorrelation time of the chain
(\citealt{goodman}), we find that the resulting chain has $\gtrsim4000$
independent posterior samples.
These samples provide an approximation to the marginalized probability
distribution for \ratepars.

@@

The detection efficiency also depends on the geometric transit probability
$R_\star/a$.
Since we are modeling the distribution in the period--radius plane, we need to
compute the transit probability marginalized over stellar radius and mass.
This marginalized distribution scales only with the period of the orbit as
$\propto \period^{-2/3}$.
In theory, this marginalization should be over the \True\ distribution of
these parameters in the selected stellar catalog but we'll approximate it by
the empirical distribution; a reasonable simplification given the size of the
dataset.
At a period of 10 days\footnote{This period is chosen arbitrarily because the
power law only needs to be normalized at one point.}, the median transit
probability in the selected sample of stars is $5.061\%$ so we model the
transit probability\footnote{We are using the letter $Q$ to indicate
probabilities since we are already using $P$ to mean period.} as a function of
period as
\begin{eqnarray}\eqlabel{transitprob}
\transitprob (\period) &=&
    0.05061\,\left[\frac{\period}{10\,\mathrm{days}}\right]^{-2/3} \quad.
\end{eqnarray}
This expression is clearly only valid for $\period \gtrsim 1.4\,\mathrm{days}$
but the dataset that we are using (\citealt{petigura}) explicitly only
includes periods longer than five days so this is not a problem.
We're using the \emph{median} transit probability (instead of the mean)
because it is a more robust estimator in the presence of outliers but in our
experiments, the results do not seem to be very sensitive to this choice.

@@

In order to get a feeling for the constraints provided by our method and to
explore any biases introduced by ignoring the observational uncertainties, we
start by ``observing'' two synthetic catalogs from qualitatively different
known occurrence rate density functions.
For each of these simulations, we take the completeness function computed by
\citet{petigura} as given.
In general, \eq{poisson-like} can be sampled using a procedure called thinning
(\citealt{poisson}) but for our purposes, we'll simply consider a piecewise
constant rate density evaluated on a fine grid in log-period and log-radius.
For this discrete function, the generative procedure is simple;
\begin{enumerate}
{\item loop over each grid cell $i$,}
{\item draw Poisson random integer $K_i\sim\mathrm{Poissson}(\obs{\rate}_i)$
with the observable rate density in the cell, and}
{\item distribute $K_i$ catalog entries in the cell randomly.}
\end{enumerate}
We then choose fractional observational uncertainties on the radii from the
\citet{petigura} catalog and apply them to the true catalog as Gaussian noise.

@@

Since no Earth analogs have been found, any constraints on this density must
be extrapolated from the existing observations.
This is generally done by assuming a functional form for the occurrence rate
density, constraining it using the observed candidates and extrapolating.
All published extrapolations are based on rigid models of the occurrence rate
density (for example, a power law) fit to the catalog and evaluated at the
location of Earth (\citealt{catanzarite, traub}).
\citet{petigura} used their catalog of planet candidates to constrain the rate
of Earth analogs in a specific period--radius bin assuming an extremely rigid
model: \emph{flat in logarithmic period}.
These results are all sensitive to the choice of extrapolation function and
the specific definition of ``Earth analog''.

@@

Figures~\figref{smooth-rate} and~\figref{simulation-rate} compare our results
and the results of the \citet{petigura} extrapolation procedure when applied
to the synthetic catalogs.
Since these catalogs were simulated from a known population model, we know the
true value of \gammaearth\ and it is indicated in the figures with a vertical
gray line.
In both cases, our method returns a less precise but more accurate result for
the rate density and the error bars given by the functional extrapolation
are overly optimistic.
One major effect that leads to this bias is that the period distribution is
not flat.
Restricting the result to only include uniform models is equivalent to
applying an extremely informative prior that doesn't have enough freedom to
capture the complexity of the problem.
As a result, the posterior constraints on \gammaearth\ are dominated by this
prior choice and the resulting uncertainties are much smaller than they should
be.

@@

The equivalent results for the radius distribution are shown in
Figures~\figref{radius} and~\figref{linear-radius}.
\Fig{radius} shows the log-radius occurrence rate density integrated over
various logarithmic bins in period.
The distributions in each period bin are qualitatively consistent; the
rate density is dominated by small planets (around two Earth radii) with
potential ``features'' near $\radius\sim3\radius_\oplus$ and $\radius\sim
10\radius_\oplus$.
These features appear in every period bin.
They were also detected---using a completely different dataset and
technique---by \citet{dong} and a similar result is visible in the occurrence
rate determined by \citet[][their Figure 7]{fressin-fp} at low
signal-to-noise.
\Fig{linear-radius} shows the same result but presented as a function of
linear radius.
In these coordinates, the rate density in a single bin is no longer
uniform; instead, scales as inverse radius.

@@

Our inferred rate density of Earth analogs (\eqalt{ge-result}) is not
consistent with the previously published result even though we use
\emph{exactly the same dataset}.
This inconsistency is due to the different assumptions made.
In particular, for their analysis, \citet{petigura} asserted that the period
distribution for small planets on long orbits is flat in logarithmic period.
If we enforce the same constraint, we reproduce their result but when we relax
this assumption, allowing the data to constrain the period distribution, we
find evidence for a \emph{decreasing} period distribution as a function of
logarithmic period (see \fig{period}).

@@

We have developed a hierarchical probabilistic framework for inferring the
population of exoplanets based on noisy incomplete catalogs.
This method incorporates systematic treatment of observational uncertainties
and detection efficiency.
One major benefit of this framework is that it provides the best possible
probabilistic measurements of the population under the assumptions listed in
\sect{intro} and repeated below.
After demonstrating the validity of our method on two qualitatively different
synthetic exoplanet catalogs, we run our inference on a published catalog of
small exoplanet candidates orbiting Sun-like stars (\citealt{petigura}) to
determine the occurrence rate density these planets as a function of period
and radius.
We extrapolate this measurement to the location of Earth and constrain the
rate density of Earth analogs with large error bars.
In order to perform this extrapolation, we don't assume a specific functional
form for the rate density.
Instead, we only assume that it is a smooth function of logarithmic period and
radius.

@@

Our extrapolation of the rate density to the location of Earth is more general
and conservative than any previously published method.
We find a rate density of Earth analogs that is inconsistent with the result
published by \citet{petigura}.
This discrepancy can be largely attributed to the rigidity of the assumptions
about the period distribution.
Our extrapolation is also less confident than previous measurements.
Again, this difference is due to the fact that we allow a much more flexible
extrapolation function.
This is another illustration that, against the standard data analysis
folklore, the correct use of flexible models is \emph{conservative}.

@@

\item {\bf Conditional independence}\quad
We assume that every object in the catalog is a conditionally independent
draw from the observable occurrence rate density.
This is a bad assumption when applying this method to a different catalog
where multiple transiting systems are included.
In practice, the best first step towards relaxing this assumption is probably
to follow \citet{tremaine} and assume that the mutual inclination distribution
is the only source of conditional dependence between planets.
For this \paper, the assumption of conditional independence is justified
because the dataset explicitly includes only systems with a single transiting
exoplanet.

@@

\item {\bf Given empirical detection efficiency}\quad
\citet{petigura} determined the end-to-end detection efficiency of their
planet detection pipeline as a function of \True\ period and radius by
injecting synthetic signals into real light curves and testing recovery.
We used these simulations as an exact representation of the detection
efficiency of the catalog but there are several missing components.
The biggest effect is probably the fact that this formulation doesn't include
the selection of only the \emph{most detectable signal} in each light curve.
This bias will be largest in the parts of parameter space where the baseline
detection efficiency is lowest: at long periods and small radius.
As a result, our inferences (and the results from \citealt{petigura}) about
the occurrence rate of small planets on long periods is probably
\emph{underestimated} relative to \Truth.
In detail there is another limitation due to the fact that the stellar
parameters are only known noisily and the transit light curve only constrains
the radius ratio.
This means that the marginalized detection efficiency should be measured as a
function of radius ratio and the interpretation in terms of \True\ radius is
only approximately correct.
Given the size of the dataset and the number of injection simulations, this
effect should be small.

@@

Under the assumptions that we have made here, our inference of the occurrence
rate density of exoplanets places a probabilistic constraint on the number of
transiting Earth analogs in the existing \kepler\ dataset.
If we adopt the definition of ``Earth-like'' from \citet[][$200 \le
\period/\mathrm{day} < 400$ and $1 \le \radius/\radius_\oplus < 2$]{petigura},
and integrate the product inferred rate density function and the geometric
transit probability (\eqalt{transitprob}) over this bin, we find that the
expected number of Earth-like exoplanets transiting the stars in the sample of
Sun-like stars chosen by \citet{petigura} is
\begin{eqnarray}\eqlabel{ntransit}
N_{\oplus,\,\mathrm{transiting}} &=& 9.2_{-4.0}^{+5.9}
\end{eqnarray}
where the uncertainties are only on the expectation value and don't include
the Poisson sampling variance.
This is an exciting result because it means that, if we can improve the
sensitivity of exoplanet search pipelines to small planets orbiting on long
periods then we should find some Earth analogs in the existing data.
Furthermore, because of the treatment of multiple transiting systems in the
catalog, the \True\ expected number of transiting Earth-like exoplanets
orbiting Sun-like stars is probably larger than the values in \eq{ntransit}!
